\Title

### 2021-03-01 latex/paper/main.tex
Slight update of notation for matrix stacking

\subsection{Notation}
We use the following notation in this article.
For \(A \in \R^{n\times m}\) and \(B \in \R^{p \times m}\),
\((A,B)\tpose \in \R^{(n+p)\times m}\) is the matrix obtained by stacking \(A\)
on top of \(B\). 
For \(T > 0\), the set of real numbers modulo \(T\) is denoted \(\Rt{T}\), with
\(\Rt{\infty} := \R\).
The gradient of the matrix-valued function 
\(A : \R^m \rightarrow \R^{n\times n}\) is the block matrix of stacked partial
derivatives, 
\(\nabla_xA := (\pdiff{A}{x_1},\ldots,\pdiff{A}{x_m})\tpose \in \R^{nm \times n}\).
Given two matrices \(A \in \R^{n \times m}\) and \(B \in \R^{r \times s}\), the
Kronecker product \cite{kronprod} \(A \otimes B \in \R^{nr \times ms}\) is the
matrix
\begin{equation}\label{eqn:kronprod}
    A \otimes B = \begin{bmatrix}
        A_{1,1}B & \cdots & A_{1,m} B \\
        \vdots & \ddots & \vdots \\
        A_{n,1} B & \cdots & A_{n,m} B
    \end{bmatrix} 
    .
\end{equation}
The Poisson bracket \cite{landau_mechanics} between the functions
\(f(q,p)\) and \(g(q,p)\) is
\begin{equation}\label{eqn:poisson-bracket}
    [f,g] := \sum \limits_{i=1}^n \pdiff{f}{p_i}\pdiff{g}{q_i} - 
        \pdiff{f}{q_i}\pdiff{g}{p_i}
    .
\end{equation}
Finally, the Kronecker delta \(\delta_i^j = 1\) if \(i = j\) and \(0\)
otherwise.

@@

\begin{lemma}\label{lemma:B-orthogonal}
    Suppose Assumption \ref{assm:B-const} holds. Then
    there exists a nonsingular matrix \(\hat{T} \in \R^{k \times k}\) 
    so that the regular feedback transformation 
    \[
        \tau = \hat{T} \hat{\tau}
    \] 
    has a new input matrix \(\hat{B}\) for \(\hat{\tau}\) which is left
    semi-orthogonal.  
    That is, \(\hat{B}\tpose \hat{B} = \Id{k}\).
\end{lemma}
\begin{proof}
    Since \(B\) is constant and full rank, it has a singular value decomposition 
    \(B = U\tpose \Sigma V\) where 
    \(\Sigma = (\diag{\sigma_1,\ldots,\sigma_k}, \Zmat{(n-k)\times k})\tpose\),
    \(\sigma_i > 0\), and \(U \in R^{n \times n}\),
    \(V \in \R^{k \times k}\) are unitary matrices \cite{calculating_svd}.
    Defining \(T = \diag{1/\sigma_1^2,\ldots,1/\sigma_n^2}\) and assigning the
    regular feedback transformation \(\tau = V T \hat{\tau}\) yields a new input
    matrix \(\hat{B} = B V T\) for \(\hat{\tau}\) such that
    \(\hat{B}\tpose \hat{B} = T\tpose \Sigma\tpose \Sigma T = \Id{k}\).
\end{proof}

In light of Lemma \ref{lemma:B-orthogonal}, there is no loss of generality in
assuming that the input matrix is left semi-orthogonal.
Let now \(\mathbf{B} \in \R^{n\times n}\) be the following matrix:
\[
    \mathbf{B} = 
    \begin{bmatrix}
        B^\perp \\
        B\tpose \\
    \end{bmatrix}
    .
\]
Since \(B^\perp\) is a left annihilator of \(B\) and both \(B^\perp\) and
\(B\tpose\) are right semi-orthogonal, it is easy to show that \(\mathbf{B}\) is
orthogonal.

\begin{thm}\label{thm:simply-actuated}
    Take the Hamiltonian system ~\eqref{eqn:hamiltonian}.
    Under Assumptions \ref{assm:B-const} and \ref{assm:B-perp},
    the coordinate transformation
    \(\left(\tilde{q} = \mathbf{B}q, \tilde{p} = \mathbf{B}p\right)\)
    is a canonical transformation.
    The resulting dynamics are given by 
    \begin{gather}\label{eqn:simple-hamiltonian}
        \mathcal{H}(\tilde{q},\tilde{p}) = 
        \frac{1}{2} \tilde{p}\tpose \Minv(\tilde{q}) \tilde{p} + V(\tilde{q})
        , \\
       \begin{cases}
           \dot{\tilde{q}} = \Minv(\tilde{q})\tilde{p}
           , \\
           \dot{\tilde{p}} = -\frac{1}{2} (\Id{n} \otimes \tilde{p}\tpose)
           \nabla_{\tilde{q}} \Minv(\tilde{q}) \tilde{p} \\
           \phantom{---} - \nabla_{\tilde{q}} V(\tilde{q}) + \simpleB \tau
            ,
        \end{cases} \nonumber
    \end{gather}
    where 
    \(\Minv(\tilde{q}) := \mathbf{B}D^{-1}(\mathbf{B}\tpose
    \tilde{q})\mathbf{B}\tpose\)
    and
    \(V(\tilde{q}) := P(\mathbf{B}\tpose \tilde{q})\).
\end{thm}
\begin{proof}
    Since \(\mathbf{B}\) is constant, this transformation satisfies
    \(\pdiff{\tilde{q}_i}{p_m} = \pdiff{\tilde{p}_i}{q_m} = 0\) for all 
    \(i,m \in \{1,\ldots,n\}\).
    Hence, \([\tilde{q}_i, \tilde{q}_j] = [\tilde{p}_i,\tilde{p}_j] = 0\).
    Then, since \(\mathbf{B}\) is orthogonal, 
    \((\mathbf{B}_i)\tpose (\mathbf{B}\tpose)_j = \delta_i^j\).
    Hence, the Poisson bracket between \(\tilde{p}_i\) and \(\tilde{q}_j\) is 
    \begin{align*}
        [\tilde{p}_i, \tilde{q}_j] &= \sum\limits_{m=1}^n A_{i,m}A_{j,m} 
                                   = (\mathbf{B}_i)\tpose (\mathbf{B}\tpose)_j
                                   = \delta_i^j
                                   .
    \end{align*}
    By (45.10) in \cite{landau_mechanics}, this is a canonical transformation,
    so the new Hamiltonian is
    \(\mathcal{H}(\mathbf{B}\tpose q, \mathbf{B}\tpose p)\).
    Finally, since \(\dot{\tilde{p}} = \mathbf{B}\tpose \dot{p}\), the input
    matrix for the system in \((\tilde{q},\tilde{p})\) coordinates is
    \[
        \mathbf{B}B = \simpleB
        ,
    \]
    which proves the theorem.
\end{proof}
These new coordinates are called 
\textit{simply actuated coordinates}.
The first \((n-k)\) coordinates, labelled \(q_u\), are named
the \textit{unactuated coordinates}; 
the remaining \(k\) coordinates, labelled \(q_a\), are the
\textit{actuated coordinates}.
The corresponding \((p_u, p_a)\) are the \textit{unactuated} and 
\textit{actuated momenta}, respectively.

@@

The nonholonomic constraint \(h(q,p)\) is \textit{virtual} because it is
enforced by the controller \(\tau(q,p)\), rather than arising from the physics
of the system.
The constraint manifold \(\Gamma\) is a \(2(n-k)\)-dimensional
closed embedded submanifold of \(\mathcal{Q} \times \R^n\).
Hence, the VNHC reduces the dimensionality of the dynamics from \(2n\) to
\(2(n-k)\). 
In particular, if \(k = (n-1)\), the constraint manifold is always
2-dimensional.

We often want to stabilize the constraint within a neighbourhood of \(\Gamma\).
To see when this is possible, let us define the error output \(e = h(q,p)\).
Notice that if any component of \(e_i\) has relative degree 1, we may not be able
to stabilize \(\Gamma\) -- we can always guarantee \(e_i \to 0\), but not
necessarily \(\dot{e}_i \to 0\).
It is for this reason that we define the following special type of VNHC.

@@

The authors of
\cite{nhvc_dynamic_walking,hybrid_zero_dynamics_bipedal_nhvcs,nhvc_incline_walking}
observed that a relation which uses only the unactuated conjugate of momentum
will have vector relative degree \(\{2,\ldots,2\}\).
In fact, this is a hard requirement: any VNHC which depends on the actuated
momentum cannot be regular. 
Before providing a characterization of regularity which proves this statement,
we define the following matrices as shorthand:
\begin{align}
    \mathcal{M}(q,p) &:= (\Id{n-k} \otimes p\tpose)\nabla_{q_u}\Minv(q) 
        ,\\
    \mathcal{D}(q,p_u) &:= dh_q(q,p_u) \Minv(q) 
    .
\end{align}

>>>>

\subsection{Notation}
We use the following notation in this article.
For \(A \in \R^{n\times m}\) and \(B \in \R^{p \times m}\),
define \([A;B] \in \R^{(n+p)\times m}\) as the matrix obtained by stacking \(A\)
on top of \(B\). 
For \(T > 0\), the set of real numbers modulo \(T\) is denoted \(\Rt{T}\), with
\(\Rt{\infty} := \R\).
The gradient of the matrix-valued function 
\(A : \R^m \rightarrow \R^{n\times n}\) is the block matrix of stacked partial
derivatives, 
\(\nabla_xA := [\pdiff{A}{x_1};\ldots;\pdiff{A}{x_m}] \in \R^{nm \times n}\).
Given two matrices \(A \in \R^{n \times m}\) and \(B \in \R^{r \times s}\), the
Kronecker product \cite{kronprod} \(A \otimes B \in \R^{nr \times ms}\) is the
matrix
\begin{equation}\label{eqn:kronprod}
    A \otimes B = \begin{bmatrix}
        A_{1,1}B & \cdots & A_{1,m} B \\
        \vdots & \ddots & \vdots \\
        A_{n,1} B & \cdots & A_{n,m} B
    \end{bmatrix} 
    .
\end{equation}
The Poisson bracket \cite{landau_mechanics} between the functions
\(f(q,p)\) and \(g(q,p)\) is
\begin{equation}\label{eqn:poisson-bracket}
    [f,g] := \sum \limits_{i=1}^n \pdiff{f}{p_i}\pdiff{g}{q_i} - 
        \pdiff{f}{q_i}\pdiff{g}{p_i}
    .
\end{equation}
Finally, the Kronecker delta \(\delta_i^j = 1\) if \(i = j\) and \(0\)
otherwise.

@@

\begin{lemma}\label{lemma:B-orthogonal}
    Suppose Assumption \ref{assm:B-const} holds. Then
    there exists a nonsingular matrix \(\hat{T} \in \R^{k \times k}\) 
    so that the regular feedback transformation 
    \[
        \tau = \hat{T} \hat{\tau}
    \] 
    has a new input matrix \(\hat{B}\) for \(\hat{\tau}\) which is left
    semi-orthogonal.  
    That is, \(\hat{B}\tpose \hat{B} = \Id{k}\).
\end{lemma}
\begin{proof}
    Since \(B\) is constant and full rank, it has a singular value decomposition 
    \(B = U\tpose \Sigma V\) where 
    \(\Sigma = [\diag{\sigma_1,\ldots,\sigma_k}; \Zmat{(n-k)\times k}]\),
    \(\sigma_i > 0\), and \(U \in R^{n \times n}\),
    \(V \in \R^{k \times k}\) are unitary matrices \cite{calculating_svd}.
    Defining \(T = \diag{1/\sigma_1^2,\ldots,1/\sigma_k^2}\) and assigning the
    regular feedback transformation \(\tau = V T \hat{\tau}\) yields a new input
    matrix \(\hat{B} = B V T\) for \(\hat{\tau}\) such that
    \(\hat{B}\tpose \hat{B} = T\tpose \Sigma\tpose \Sigma T = \Id{k}\).
\end{proof}

In light of Lemma \ref{lemma:B-orthogonal}, there is no loss of generality in
assuming that the input matrix is left semi-orthogonal.
Let now \(\mathbf{B} = [B^\perp; B^\tpose]\).
Since \(B^\perp\) is a left annihilator of \(B\) and both \(B^\perp\) and
\(B\tpose\) are right semi-orthogonal, it is easy to show that \(\mathbf{B}\) is
orthogonal.

\begin{thm}\label{thm:simply-actuated}
    Take the Hamiltonian system ~\eqref{eqn:hamiltonian}.
    Under Assumptions \ref{assm:B-const} and \ref{assm:B-perp},
    the coordinate transformation
    \(\left(\tilde{q} = \mathbf{B}q, \tilde{p} = \mathbf{B}p\right)\)
    is a canonical transformation.
    The resulting dynamics are given by 
    \begin{gather}\label{eqn:simple-hamiltonian}
        \mathcal{H}(\tilde{q},\tilde{p}) = 
        \frac{1}{2} \tilde{p}\tpose \Minv(\tilde{q}) \tilde{p} + V(\tilde{q})
        , \\
       \begin{cases}
           \dot{\tilde{q}} = \Minv(\tilde{q})\tilde{p}
           , \\
           \dot{\tilde{p}} = -\frac{1}{2} (\Id{n} \otimes \tilde{p}\tpose)
           \nabla_{\tilde{q}} \Minv(\tilde{q}) \tilde{p} \\
           \phantom{---} - \nabla_{\tilde{q}} V(\tilde{q}) + \simpleB \tau
            ,
        \end{cases} \nonumber
    \end{gather}
    where 
    \(\Minv(\tilde{q}) := 
    \mathbf{B}D^{-1}(\mathbf{B}\tpose \tilde{q})\mathbf{B}\tpose\)
    and
    \(V(\tilde{q}) := P(\mathbf{B}\tpose \tilde{q})\).
\end{thm}
\begin{proof}
    Since \(\mathbf{B}\) is constant, this transformation satisfies
    \(\pdiff{\tilde{q}_i}{p_m} = \pdiff{\tilde{p}_i}{q_m} = 0\) for all 
    \(i,m \in \{1,\ldots,n\}\).
    This implies the Poisson brackets \([\tilde{q}_i, \tilde{q}_j]\)
    and \([\tilde{p}_i,\tilde{p}_j]\) are both zero.
    Then, since \(\mathbf{B}\) is orthogonal, 
    \((\mathbf{B}_i)\tpose (\mathbf{B}\tpose)_j = \delta_i^j\).
    Hence, 
    \begin{equation*}
        [\tilde{p}_i, \tilde{q}_j]
            = \sum\limits_{m=1}^n \mathbf{B}_{i,m}\mathbf{B}_{j,m} 
            = (\mathbf{B}_i)\tpose (\mathbf{B}\tpose)_j
            = \delta_i^j
            .
    \end{equation*}
    By (45.10) in \cite{landau_mechanics}, this is a canonical transformation
    with new Hamiltonian
    \(\mathcal{H}(\mathbf{B}\tpose q, \mathbf{B}\tpose p)\).
    Finally, since \(\dot{\tilde{p}} = \mathbf{B}\tpose \dot{p}\), the input
    matrix for the system in \((\tilde{q},\tilde{p})\) coordinates is
    \[
        \mathbf{B}B = \simpleB
        ,
    \]
    which proves the theorem.
\end{proof}
These new coordinates are called 
\textit{simply actuated coordinates}.
The first \((n-k)\) coordinates, labelled \(q_u\), are named
the \textit{unactuated coordinates}; 
the remaining \(k\) coordinates, labelled \(q_a\), are the
\textit{actuated coordinates}.
The corresponding \((p_u, p_a)\) are the \textit{unactuated} and 
\textit{actuated momenta}, respectively.

@@

The nonholonomic constraint \(h(q,p)\) is called \textit{virtual} because it is
enforced by the controller \(\tau(q,p)\), rather than being a constraint which
arises from the physics of the system.
The constraint manifold \(\Gamma\) is a \(2(n-k)\)-dimensional
closed embedded submanifold of \(\mathcal{Q} \times \R^n\).
Hence, the VNHC reduces the dimensionality of the dynamics from \(2n\) to
\(2(n-k)\). 
In particular, if \(k = (n-1)\), the constraint manifold is always
2-dimensional.

We often want to stabilize the constraint within a neighbourhood of \(\Gamma\).
To see when this is possible, let us define the error output \(e = h(q,p)\).
If any component of \(e_i\) has relative degree 1, we may not be able
to stabilize \(\Gamma\) -- we can always guarantee \(e_i \to 0\), but not
necessarily \(\dot{e}_i \to 0\).
It is for this reason that we define the following special type of VNHC.

@@

The authors of
\cite{nhvc_dynamic_walking,hybrid_zero_dynamics_bipedal_nhvcs,nhvc_incline_walking}
observed that only relations which use the unactuated conjugate of momentum
have vector relative degree \(\{2,\ldots,2\}\).
However, this is not enough to guarantee the

We now provide a computational characterization of regularity
we define the following matrices as shorthand:
\begin{align}
    \mathcal{M}(q,p) &:= (\Id{n-k} \otimes p\tpose)\nabla_{q_u}\Minv(q) 
        ,\\
    \mathcal{D}(q,p_u) &:= dh_q(q,p_u) \Minv(q) 
    .
\end{align}
